---
title: "Election 2020 Simulation (Uninformative vs. Informative Prior)"
author: "Mahmoud Abu Ghzalah"
date: "2/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Did Trump have more support than Biden in 2020?

First use an uninformative prior, here set all shape parameters in the dirichlet distrubtion = 1

Data is sourced from a 2020 Qunnipiac University National poll of 1426 self identified likely voters

Link: https://poll.qu.edu/images/polling/us/us10222020_udxf16.pdf

```{r echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(gtools)
```

Create a table to store the proportions of voters

```{r}
(voter_prop <- tibble("Biden" = 0.51, "Trump" = 0.41, 
                     "Other" = 0.02, "No_Opinion" = 0.06))
```

Create a table to store the number of likely voters in each category

```{r}
# Variable to store number of respondents to poll
voter_sample <- 1426
```



Multiply sample size by proportions to get counts for each category

```{r}
(voter_count <- voter_sample * voter_prop) 
```

Voters in each category are not integers. I'll round them to the nearest whole number

```{r}
(voter_count <- map_dbl(voter_count, round, digits = 0))
```


We have an extra observation compared to the sample size, not sure what the implication of this would be. 

If there's a more elegant solution, let me know. 

Use the vector of voter counts to simulate 10,000 draws from the posterior Dirichlet distribution with an uninformative prior

```{r}
# Set seed for reproducibility
set.seed(12345)

# Simulate 10,000 draws from posterior distribution
sim_1 <- rdirichlet(10000, alpha = voter_count)
head(sim_1)
```

What is the probability Trump had more support than Biden? ie: $\alpha_2 + y_2 > \alpha_1 + y_1$

```{r}
hypo_1 <- sim_1[, 2] - sim_1[, 1]
```

# Plot distribution of the hypothesis:

```{r fig.align='center'}
plot(ecdf(hypo_1), xlab = 'Vote Gap (proportion)', main = "Trump vs. Biden (Uninformative Prior)")
```

It's pretty unlikely that Trump had more support than Biden according to our simulation

Find the probability Trump had more support than Biden by counting the proportion of elements in hypo_1 that are greater than 0, this is the number of simulated elections in which Trump had more support than Biden

```{r}
length(hypo_1[hypo_1 > 0]) / length(hypo_1)
```

There wasn't a single simulation using the uninformative prior in which Trump would have had more support than Biden

I'll now use an informative prior - we'll adjust the shape parameters of the prior to reflect voter preferences from the 2016 Trump v. Clinton election

I begin by creating a table to store voter preferences from the 2016 election

Using data sourced from the Federal Elections Committee: https://www.fec.gov/documents/1889/federalelections2016.pdf

Trump received 46.09% of the popular vote

Clinton received 48.18% of the popular vote

```{r}
informative_prior <- tibble("Trump" = 0.4609, 
                            "Clinton" = 0.488, "Other" = 5.73)
```


I adjust the number of categories from the 2020 election poll by combining "No opinion" and "other" into one category

```{r}
voter_prop <- voter_prop %>% mutate(
  "Neither" = No_Opinion + Other) %>% 
  select(-c(No_Opinion, Other))

voter_prop
```


```{r}
# Combine categories in the voter_count vector 
voter_count <- c(voter_count[1:2], voter_count[3] + voter_count[4])
```

Apply the informative prior to the number of likely voters to calculate alphas (shape parameters of the Dirichlet posterior)

```{r}
(alpha_params <- informative_prior * voter_sample)
```

Create vector that contains parameters of posterior Dirichlet distribution by summing alpha parameter and 2020 poll responses

```{r}
# Convert post_params from a tibble to a vector
(post_params <- (alpha_params + voter_count) %>% slice(1) %>% as.numeric())
```



Simulate 10,000 draws from the posterior distribution using the informative prior

```{r}
sim_2 <- rdirichlet(10000, alpha = post_params)
head(sim_2)
```

What is the probability Trump had more support than Biden? 

ie: $\alpha_2 + y_2 > \alpha_1 + y_1$

```{r}
hypo_2 <- sim_2[, 2] - sim_2[, 1]
```

Plot cumulative distribution of the vote gap between Biden and Trump

```{r fig.align= 'center'}
plot(ecdf(hypo_2), xlab = 'Vote Gap (proportion)', main = "Trump vs. Biden (Informative Prior)")
```



Number of simulated elections where Trump wins more votes than Biden

```{r}
length(hypo_2[hypo_2 > 0])

# Find porportion to obtain probability

length(hypo_2[hypo_2 > 0]) / length(hypo_2)
```

The probability that Trump secures more votes than Biden among likely voters is now 2.39% using the informative prior

Distribution of the predicted vote gap using informative prior:

```{r fig.align='center'}
hypo_2 %>% tibble() %>% ggplot() +
  geom_histogram(aes(x = hypo_2), bins = 60) +
  labs(x = "Vote Gap (Proportion)", y = "Frequency", title = "Distribution of Vote Gap (Informative Prior)")
```

Distribution of the predicted vote gap using an uninformative prior:

```{r fig.align='center'}
hypo_1 %>% tibble() %>% ggplot() +
  geom_histogram(aes(x = hypo_1), bins = 60) +
  labs(x = "Vote Gap (Proportion)", y = "Frequency", title = "Distribution of Vote Gap (Uninformative Prior)")
```

Overlay the two distributions using a frequency polygon

```{r fig.align='center'}
ggplot() +
  geom_freqpoly(data = tibble(hypo_1), aes(x = hypo_1, colour = "Uninformative Prior"), bins = 100) +
  geom_freqpoly(data = tibble(hypo_2), aes(x = hypo_2, colour = "Informative Prior"), bins = 100) +
  labs(color = "Legend text", title = "Posterior Distribution: Informative vs. Uninformative Prior", x = "Vote Gap (Proportion)")
```




